---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE, 
  warning = FALSE,
  comment = "#>"
  #, fig.path = "man/figures/README-"
)
```

# RstoxStrata
**Define strata for survey indices. R package, updated `r packageDate("RstoxStrata")`.**

This package helps to define strata from bathymetric grids for survey indices. The survey indices can be calculated using the [StoX software](https://stoxproject.github.io/) but also manually outside StoX. See [the user guide](https://deepwaterimr.github.io/RstoxStrata/articles/RstoxStrata.html) for further details.

## Installation

The package can be installed from GitHub using the [**devtools**](https://cran.r-project.org/web/packages/devtools/index.html) (or [remotes](https://cran.r-project.org/web/packages/remotes/index.html)) package:

```{r eval = FALSE}
devtools::install_github("DeepWaterIMR/RstoxStrata")
```

## Usage

See [the user guide](https://deepwaterimr.github.io/RstoxStrata/articles/RstoxStrata.html) and [gallery of strata systems](https://deepwaterimr.github.io/RstoxStrata/articles/Examples.html) to learn how to use the package.

## Citation

If you use the package for stock assessment or scientific articles, please cite it:

```{r}
citation("RstoxStrata")
```

