---
title: "Estimation of strata with RstoxStrata"
author: "Mikko Vihtakari (Institute of Marine Research)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  bookdown::html_document2:
    toc: true
    toc_float: true
    code_folding: show
    number_sections: false
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{RstoxStrata}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
library(knitr)
library(kableExtra)

knitr::opts_chunk$set(collapse = TRUE, message = FALSE, warning = FALSE, comment = "#>")
```

Packages required to run the examples:

```{r}
library(sf)
library(dplyr)
library(tidyr)
library(tibble)
library(RstoxStrata)
library(ggOceanMaps)
library(cowplot)
```

```{r, include = FALSE}
# Official EggaN index for comparison. Acquired from strata-and-survey-indices.Rmd

official <- structure(list(year = c(1996L, 1997L, 1998L, 1999L, 2000L, 2001L, 
                                    2002L, 2003L, 2004L, 2005L, 2006L, 2007L, 2008L, 2009L, 2011L, 
                                    2013L, 2015L, 2017L, 2019L), value = c(81.799, 82.345, 82.717, 
                                                                           104.194, 80.587, 94.785, 97.607, 109.489, 103.62, 86.944, 81.839, 
                                                                           88.257, 84.909, 108.876, 84.968, 63.185, 78.013, 59.382, 46.967
                                    )), class = c("tbl_df", "tbl", "data.frame"), row.names = c(NA, 
                                                                                                -19L))
```


# Introduction

Fisheries assessment models use catch data and population trends to estimate the fishing quota (Total Allowable Catch, TAC), i.e., the safe or optimal biomass to remove from a population to ensure population continuity (sustainability). While catch data primarily govern the absolute levels in TAC estimates, population trends dictate the relative changes in TAC from year to year and help the assessment model differentiate fishing effort- and efficiency-related catch variations from actual population fluctuations. Therefore, estimating population trends is integral to fisheries assessment and can have considerable economic and ecologic consequences.  

Estimating population trends requires deciphering how much fish there is in the sea, commonly done using survey indices tailored for a particular species, region, and sampling setup. In essence, these survey indices are estimated by summing up numbers or weights of fish within areas called strata. The strata model distribution of a stock within a particular region of interest - often delimited by political or population boundaries. Most survey indices use multiple strata, assuming that the abundance of a species, population structure, or sampling setup considerably differs within the region of interest. The formula for estimating the number (or weight) of fish within a stratum for gear that covers an area, such as trawls, multiplies an average number of fish per areal unit (=density) by area of the stratum. In other words, the density of fish in each stratum is weighted by the area. Since the area is a multiplier, it considerably impacts the outcome. Imprecision in the areal estimation of strata or missing samples within a stratum may greatly affect the estimated population trends and hence TAC. 

Depth is among the best factors to describe the distribution of fish species [REFS]. Consequently, bottom depth intervals are commonly incorporated into strata to model the vertical distribution of a species. Using bottom depth can lead to imprecision in areal estimation for species that inhabit narrow depth ranges along the continental slopes, such as the Greenland halibut. Imprecision in the areal estimation stemming from poorly charted bathymetry and low or inconsistent model resolution can be an order of magnitude and directly proportional to survey index estimates. 

Here we discuss pitfalls in making strata for survey indices and study the consequences of filtering disconnected regions from strata using synthetic populations and Northeast Arctic Greenland halibut as example data. We suggest a transparent framework to estimate strata from bathymetric grid data programmed to an R package called RstoxStrata, which is integrated into the StoX software for survey index calculation. While the framework is applied to bottom depth, it can be used for any gridded spatial data such as surface temperature, salinity, or modeled prey density. 

# Material and methods

The proposed way of defining strata here attempts to simplify the strata making. Further, the process is transparent and reproducible. The distribution of a species is modeled using two components: 1) geographic regions (rectangles, called geostrata) are used to model the spatial distribution, and 2) bottom depth intervals within the geographic regions are used as a proxy of the vertical distribution. The depth data are acquired from the highest openly available evenly gridded **depth model** at the time of writing, the [General Bathymetric Chart of the Oceans](https://www.gebco.net/data_and_products/gridded_bathymetry_data/) 15-arcsecond 2021 grid (approximately 1/4 nautical miles in the latitude direction). The *GEBCO_2021 Grid (ice surface elevation)* netCDF file is downloaded (4 Gb) and placed in a central location on the computer:

```{r}
link <- "~/Desktop/gebco_2021/GEBCO_2021.nc"
```


We use the "EggaNord" survey in the Northeast Atlantic spawning grounds of Greenland halibut as an example. The survey and its strata system are used to estimate the primary stock index for [Northeast Arctic Greenland halibut](https://www.ices.dk/sites/pub/Publication%20Reports/Advice/2021/2021/ghl.27.1-2.pdf). The original system assigns stations to strata based on latitude and gear depth [@Høines2008], while the StoX software uses position only in the assignment [@Johnsen2019] (Table \@ref(tab:origareas)). The calculus for the strata area used in the original system has been lost and is not reproducible. Since the new StoX based approach uses solely position in station allocation to strata, we need to define the longitude limits as well to mimic the old strata system.

```{r origareas, echo=FALSE, message=FALSE, warning=FALSE}
data(eggan_ghl_original_strata)

x <- eggan_ghl_original_strata
x$lat.interval <- paste(x$lat.min, x$lat.max, sep = "-")
names(x)[names(x) == "interval"] <- "depth.interval"
x <- na.omit(x)

x$lat.interval <- factor(x$lat.interval, levels = unique(x$lat.interval))
x$depth.interval <- factor(x$depth.interval, levels = unique(x$depth.interval))

tab <- pivot_wider(x, id_cols = "depth.interval", names_from = "lat.interval", values_from = "orig.nm2")
names(tab)[names(tab) == "lat.interval"] <- ""

kable(tab, caption = "Previously used strata system definition for NEA Greenland halibut. Latitude intervals given are along rows, depth intervals along columns and the numbers represent strata areas in square nautical miles.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

## Geostrata

The geographic rectangles, called "**geostrata**", are defined as a data frame for the RstoxStrata package. Optionally, the first column of the data frame can define the names of the geostrata. The four following numeric columns are required and define the minimum longitude, maximum longitude, minimum latitude and maximum latitude for the rectangles, respectively (Figure \@ref(fig:geostrata)). The coordinates are defined in decimal degrees (EPSG:4326). 

```{r geostrata, fig.cap = "A geostrata system attempting to mimic the EggaNord survey strata for Northeast Greenland halibut. The currently uses system uses latitude and gear depth for station allocation while the new StoX based approach uses position only."}
geostrata.df <- data.frame(
  lon.min = c(3, 10, 10, 8),
  lon.max = c(16, 17.5, 17.5, 17.5),
  lat.min = c(76, 73.5, 70.5, 68),
  lat.max = c(80, 76, 73.5, 70.5)
)

plotGeostrata(geostrata.df)
```

## Depth strata

The **depth strata** are defined as a numeric vector specifying the cut points for the depth intervals.

```{r}
depths.vec <- c(400, 500, 700, 1000, 1500)
```

## Strata estimation

Since the GEBCO bathymetry model covers the entire world, it can be cut to the region of interest using the **boundary** argument. This argument accepts vectors with four numbers in the same order than geostrata (minimum and maximum longitude, minimum and maximum latitude). 

```{r}
boundary.vec <- c(0, 17.5, 68, 80)
```

These information can then be input into the [`strataPolygon()`](https://deepwaterimr.github.io/RstoxStrata/reference/strataPolygon.html) function which performs the strata estimation. The function returns a `strataPolygon` object, which is a list containing **strata** and **geostrata** [sf](https://cran.r-project.org/web/packages/sf/index.html) polygon objects. 

```{r, stratawithfrags, fig.cap = "Initial strata system for the EggaNord survey index containing fragments that increase the strata area", results='hide'}
pols <- strataPolygon(
  bathy = link, 
  depths = depths.vec, 
  boundary = boundary.vec,
  geostrata = geostrata.df
)

plot(pols)
```

The depth strata 400-500 m between geostrata B and C is large and exceeds regularly visited stations. The [`strataPolygon()`](https://deepwaterimr.github.io/RstoxStrata/reference/strataPolygon.html) function accepts polygons as the `boundary` argument allowing manual tuning of the strata (Figures \@ref(fig:boundary) and \@ref(fig:strata)).

```{r, boundary, fig.cap = "Spatial polygons can be used as the boundary argument within strataPolygons() to manually restrict the strata extend."}
boundary <- sf::read_sf("../inst/shapes/boundary_shape.shp")
# sf::read_sf(system.file("shapes", "boundary_shape.shp", package = "RstoxStrata"))
ggOceanMaps::qmap(boundary)
```

The initial strata system contains **fragments** from sea mounts and fjords. These fragments bias the area of strata and can be removed using the `fragment.area` argument which accepts a single threshold in square kilometers. Fragments smaller than the threshold will be removed. The process of finding the correct threshold is iterative.

```{r, strata, fig.cap = "Final strata system for the EggaNord survey to estimate a survey index for Northeast Arctic Greenland halibut. The strata system attemps to mimic the system used in the assessment that bases on latitude and gear depth.", results='hide', fig.height=14}

pols <- strataPolygon(
  bathy = link, 
  depths = depths.vec, 
  boundary = boundary,
  geostrata = geostrata.df,
  fragment.area = 400
)

plot(pols, facetted = TRUE)
```

## Survey index

```{r, class.source = 'fold-hide'}
data(eggan_ghl)

stn <- eggan_ghl %>%
  mutate(depth.interval =
           cut(.$bottomdepthstart, depths.vec,
               labels = paste(depths.vec[1:(length(depths.vec)-1)],
                              depths.vec[2:(length(depths.vec))], sep = "-"))
  ) %>%
  st_as_sf(coords = c("longitudestart", "latitudestart"), crs = 4326, remove = FALSE) %>%
  rownames_to_column("id") %>%
  st_join(pols$strata[, c("interval", "geostrata.name", "area.nm2")]) %>%
  filter(!duplicated(id)) %>%
  st_join(pols$geostrata[, c("geostrata.name")] %>%
            rename("geostrata.polygon" = "geostrata.name")) %>%
  filter(!duplicated(id)) %>%
  st_set_geometry(NULL) %>%
  mutate(area.nm2 = as.numeric(area.nm2)) %>%
  left_join(.,
            eggan_ghl_original_strata %>%
              dplyr::select(geostrata.name, interval, orig.nm2) %>%
              rename("geostrata.polygon" = "geostrata.name", "depth.interval" = "interval")
  ) %>%
  rename("area.new" = "area.nm2", "area.old" = "orig.nm2") %>%
  relocate(distance, biomass, density, .after = last_col()) %>%
  suppressMessages()

old_area_old_assig <- stn %>%
  filter(!is.na(area.old)) %>%
  dplyr::select(startyear, geostrata.polygon, depth.interval, area.old, density) %>%
  rename("geostrata.name" = "geostrata.polygon",
         "interval" = "depth.interval",
         "area" = "area.old") %>%
  group_by(startyear, geostrata.name, interval, area) %>%
  summarise(mean = mean(density), sd = sd(density), n = n()) %>%
  mutate(index = area*mean/1e6,
         index.sd = area*sd/1e6,
         ci.min.t = index - qt(1 - 0.025, (n - 1))*index.sd/sqrt(n),
         ci.max.t = index + qt(1 - 0.025, (n - 1))*index.sd/sqrt(n)) %>%
  suppressWarnings() %>%
  ungroup() %>%
  group_by(startyear) %>%
  summarise(reconst = sum(index))

surv_diff <- left_join(official, old_area_old_assig %>% rename("year" = "startyear")) %>%
  mutate(diff = value - reconst)

removed_stn <- stn %>% filter(is.na(area.new) | is.na(area.old)) %>%
  mutate(reason = ifelse(is.na(depth.interval) & is.na(geostrata.polygon), "Both",
                         ifelse(is.na(depth.interval), "Depth",
                                ifelse(is.na(geostrata.polygon), "Outside",
                                       ifelse(is.na(interval), "Position" ,"Other")))))

```

# Results

## Difference in area

The reconstructed strata areas were generally lower than the original except for the deepest 1000-1500 strata, the manually adjusted strata C 400-500, and geostrata A excluding the shallowest 400-500 strata  (Figure \@ref(fig:areadiff), Table \@ref(tab:areadifftab)). There are no obvious reasons for the difference but since the method for acquiring the original strata areas is lost and over a decade old, the reconstructed areas are likely more trustworthy than the old ones. 

```{r, areadiff, class.source = 'fold-hide', fig.cap = "Difference between the original and reconstructed strata area. A) The differences are shown as a scatter plot. Color refers to depth strata and letter to geostrata name. B) The differences are shown on a map. Color refers to percentage difference. Negative percentages mean larger original area than reconstructed area.", fig.height = 7.5}

x <- dplyr::full_join(
  eggan_ghl_original_strata %>% 
    dplyr::select(geostrata.name, interval, orig.nm2),
  pols$strata %>% 
    dplyr::select(geostrata.name, interval, area.nm2) %>% 
    dplyr::mutate(area.nm2 = as.numeric(area.nm2))
) %>% 
  mutate(pr.diff = 100*(area.nm2-orig.nm2)/orig.nm2, .before = "geometry") %>% 
  st_as_sf() %>% 
  mutate(interval = factor(interval, levels = unique(.$interval)))

p1 <- ggplot(x, aes(x = orig.nm2, y = area.nm2, color = interval, label = geostrata.name)) +
  geom_abline(slope = 1, intercept = 1) +
  geom_text() +
  labs(x = "Original area (km2)",
       y = "Polygon area (km2)",
       color = "Depth strata") +
  guides(color=guide_legend(nrow=2,byrow=TRUE)) +
  theme_bw() +
  theme(legend.position = "bottom")

p2 <- basemap(x) + 
  geom_sf(data = x, aes(fill = pr.diff)) + 
  scale_fill_gradient2("Percentage\ndifference") +
  geom_sf(data = pols$geostrata,
          fill = NA, color = "black") +
  geom_sf_text(data = sf::st_centroid(pols$geostrata) %>% suppressWarnings(),
               aes(label = geostrata.name),
               fontface = 2)  +
  theme(legend.position = "bottom")

cowplot::plot_grid(p1, p2, labels = "AUTO")
```

```{r areadifftab, echo=FALSE, message=FALSE, warning=FALSE}
tab <- x %>% 
  mutate(pr.diff = round(pr.diff, 1)) %>% 
  st_drop_geometry() %>% 
  pivot_wider(., id_cols = "interval", names_from = "geostrata.name", values_from = "pr.diff")

names(tab)[names(tab) == "interval"] <- ""

kable(tab, caption = "Percentage difference between original and reconstructed strata areas. Negative percentages mean larger original area than reconstructed area.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

## Station assignment difference

The use of position based station assignment to strata led to removal of `r removed_stn %>% filter(reason == "Position") %>% nrow()` stations that were included in the old depth-latitude based assignment (Figure \@ref(fig:removalmap)). 

```{r, removalmap, class.source = 'fold-hide', fig.cap = "Position of EggaN survey stations removed from survey index calculations. The color indicates the reason for removal: depth = the recorded bottom depth is outside the depth strata, outside = position of the station is outside the geostrata, position = position of the station is outside the strata due to impression in bathymetry model and strata estimation or due to wrongly recorded coordinates. Grey dots indicate stations included in survey index calculations. The figure contains data from all years (1994-2021)."}
p <- basemap(data = removed_stn) +
  geom_sf(data = pols$strata %>% sf::st_union(), fill = NA, color = "black") +
  geom_spatial_point(data = stn, aes(x = longitudestart, y = latitudestart), size = 0.5, color = "grey") +
  geom_spatial_point(data = removed_stn, aes(x = longitudestart, y = latitudestart, color = reason)) +
  labs(color = "Reason")

reorder_layers(p)
```

Percent agreement between old and new methods figure (geostrata, depth)

Depth interval difference figure (how they get assigned)

## Difference in survey index

The reconstructed survey index using depth- and latitude-based station assignment to strata and the original strata areas had a correlation of `r round(cor(surv_diff$value, surv_diff$reconst), 3)` with the official EggaN survey index used in the assessment model. The average annual difference between these indices was `r round(mean(surv_diff$diff)*1000, 0)` tons and standard deviation `r round(sd(surv_diff$diff)*1000, 0)` tons. While these differences may sound large, for comparison purposes the official and reconstructed survey indices were practically identical. The difference is likely caused by a few stations being included/removed from survey index calculations differently and highlights the differences weighing by the area can cause to survey indices. Consequently, only the reconstructed survey indices are used in the comparison for consistency. 

Difference between old and new area, depth based and position based assignment figure

Impact of geo- and depth strata on the index figure

Correlation between geo- and depth strata figure

# Discussion
